---
title: "Downloading and cleaning Monitoring Trends in Burn Severity (MTBS) fire data"
output: github_document
---

## Requirements

Downloading and processing the data requires that you have 
[R](https://www.r-project.org/) installed, and the following R packages:

- dplyr
- rgdal
- rgeos
- tidyr

If you do not have these packages, you can try to install them with 
`install.packages()`, e.g., 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(printr)
```


```{r, eval = FALSE}
install.packages('tidyr')
```

The script `get-mtbs-data.R` pulls data from the internet, so you'll need an 
internet connection. 


## Running the script

To run the R script, you can use `Rscript` from the terminal:

```{bash, results = 'hide'}
Rscript get-mtbs-data.R
```

Or, if you prefer, you can open the file in R or Rstudio and run it interactively. 


## What it produces

The script will produce a directory with the following structure:

```
data/
  - processed/
    - fire-events.csv
    - fire-summary.csv
  - raw/
    - mtbs_fod_pts_data/
      - ...
    - us_eco_l3/
      - ...
```

### Event data

The file `fire-events.csv` has the raw data from MTBS where each row is a fire
event:

```{r, echo = FALSE}
d <- read.csv(file.path('data', 'processed', 'fire-events.csv'))
dplyr::tbl_df(d)[1:5, ]
```

The columns are as follows: 

- `fire_id`: unique event identifier
- `year`, `month`, `day`: year, month, day of ignition
- `lat`, `long`: lat and long of ignition
- `acres_burned`: number of acres burned
- `na_l3name`, `na_l2name`, `na_l1name`: level 3, 2, and 1 ecoregion names
- `yearmonth`: year and month concatenated (e.g., '2010-01')

Both of these files have preprocessed [MACA](http://maca.northwestknowledge.net/) climate data summarized at the level of month and ecoregion, 
with four variables: mean daily potential evapotranspiration 
(`mean_potential_et`), total precipitation (`total_precip`), mean daily maximum 
air temperature (`max_air_temp`), and mean daily wind speed (`mean_wind_speed`). 


### Summary data

The file `fire-summary.csv` has the number of fires and total burn area 
summarized for each of the 85 EPA level 3 ecoregions and 382 months over the 
range of dates in the MTBS data (for a total of 382 X 85 = 32470 rows), where
each row is an ecoregion by month combination. 

```{r, echo = FALSE}
d <- read.csv(file.path('data', 'processed', 'fire-summary.csv'))
dplyr::tbl_df(d)[1:5, ]
```



### Computing polygon centroids and distance matrices


```{r, message = FALSE}
library(rgdal)
library(rgeos)

# load ecoregion shapefile
ecoregion_file <- list.files(pattern = 'us_eco_l3.shp$',
                             recursive = TRUE,
                             full.names = TRUE)
ecoregions <- readOGR(ecoregion_file)
```

#### Computing centroids

Next, we'll use the `gCentroid` function from the `rgeos` package to compute 
centroids for each polygon in the ecoregion shapefile:

```{r}
centroids <- gCentroid(ecoregions, byid = TRUE)
```

Last, because some ecoregions are discontinuous, a subset of our ecoregions now
have multiple centroids. To get just one centroid per ecoregion, we can average 
across all of the centroids for those ecoregions with multiple centroids:

```{r, message=FALSE}
library(dplyr)

ecoregion_centroids <- as.data.frame(centroids) %>%
  tbl_df %>%
  mutate(na_l3name = ecoregions$NA_L3NAME) %>%
  group_by(na_l3name) %>%
  summarize(x = mean(x), 
            y = mean(y))
```

This will return a data frame with the centroids for each ecoregion. As a final 
step, you may need to compute a distance matrix, which provides pairwise 
distances among all centroids. 

#### Computing a distance matrix

```{r}
D <- ecoregion_centroids %>%
  select(x, y) %>%
  dist %>%
  as.matrix

# associate ecoregions with rows and columns
rownames(D) <- ecoregion_centroids$na_l3name
colnames(D) <- ecoregion_centroids$na_l3name

# preview part of D
D[1:3, 1:3]
```